Unsupervised Learning for Identifying Events in Active Target Experiments

### 2021-03-11 reviewer_responses.tex

we are much indebted for the positive feedback from the reviewer and your willingness to consider our manuscript for publication.
We have tried as faithfully as possible to answer the referee and we hope the revised version meets your standards for publications in NIMA.\newline
Our answers to the referee are included below here. The revised version is included as well.\newline
Yours sincerely,\newline
Daniel Bazin, Morten Hjorth-Jensen, Michelle Kuchera, Robert Solli, and Ryan Strauss

\section*{Answers to the referee}

We are very much indebted for the positive comments on our manuscript and we would like to  thank the reviewer for a careful read of our work. We have tried as faithfully as possible to answer the various comments. We have included them here for the sake of completeness, together with our responses
(highlighted in blue). We hope our responses  aid in improving the quality of the manuscript and thank the reviewer again for her/his willingness for considering our manuscript again.

\begin{itemize}
\item Reviewer comment: Line 29, Simulated data? Line 193 simulated events… from a Ar(p,p) experiment. Are these
simulation data or experimental data? The use of word simulated is confusing. Based on the context,
it appears that only experimental data is used for this work.\\
\textcolor{blue}{Our answer: We apologize for the confusion. Both simulated events and records from the AT-TPC were studied. Simulated events were generated from an event simulator written for the purpose of analysing the Ar(p,p) reaction. We added a sentence in the abstract to clarify the circumstance. }
\item Reviewer comment: Line 124-125, simulated via an integration, which is numerically costly. Rephrase and clarify how
costly quantitatively. Similar clarifications needed in Line 131, Line 145 and Table 2.\\
\textcolor{blue}{Our answer: More quantitative information about the computational cost of traditional analysis methods was added in the paragraphs mentioned by the referee.}
\item Reviewer comment: Figure 1. Add some discussions about the experimental information and physics behind the
observed patterns. For example, which direction is the incoming beam? Beam energy? Vertex at
(0,0)? The direction of the magnetic field?\\

@@

\textcolor{blue}{Our answer: Additional text has been added in the figure caption to better explain the experimental conditions of the displayed data.}
\item Reviewer comment: Line 306, Sample $\hat{x}$, presumably each of these is a 2D image x-y? what are the sample
dimensions?

\textcolor{blue}{Our answer: The text has been modified to clarify that the samples are images encoded as a square matrix with 128 elements along one axis.}

\item Reviewer comment: Lines 330 \& 331, define Ground truth labels y and predictions $\hat{y}$, how are they derived from
the raw data (2D projections of the AT-TPC)? Or the latent layer outputs of a pre-trained neural
network? Or else?
\textcolor{blue}{Our answer: We have added more text in each subsection of Section 3: Data Preparation was added to indicate where the ground truth labels were derived from. In fact, we used this data set {\it because} we had rare access to ground truth labels for experimental data.}
\item Reviewer comment: Figures 4, Figure 5, Figure 6, Figure 9, Figure 10, are data science metric results. Needs update or
move some of them to the appendix.
\textcolor{blue}{Our answer: We have followed the reviewer's suggestions and have moved the confusion matrix figures to the appendix, and we added text to clarify their meaning.}
\item Reviewer comment: Figure 7, Figure 8, elucidate the physics and understanding behind the pattern would be helpful.
Similar to the comments to Figure 1 above.
\textcolor{blue}{Our answer: These figures are now the new Figures 4 and 5. Since this is unsupervised clustering, it is normally difficult to  conjecture a clear   physical meaning behind the separation. However, the plots visualize a sampling of proton events that were attributed to a noisy and pure cluster (each cluster), allowing a reader to infer the behavior of the unsupervised learning on these events. The figure caption texts have been updated to clarify this point. }
\end{itemize}

### 2021-03-11 doc/Manuscript/article.tex
Updates from Overleaf

\begin{abstract}
% Modified by DB (6/8/2020)
%Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant. 
%on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. 
%On the semi-supervised task, we find that a logistic regression classifier trained on small labeled subsets of the latent space of these models perform very well, both on simulated and real data. On the clustering task, 
%With autoencoder neural networks we find improved descriptions of data from experiments.
This article presents novel applications of unsupervised machine learning methods to the problem of event separation in an active target detector, the Active-Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. 
The overarching goal is to group similar events in the early stages of the data analysis, thereby improving efficiency by limiting the computationally expensive processing of unnecessary events.
The application of unsupervised clustering algorithms to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on $^{46}$Ar is introduced. We explore the performance of autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network.
We find that a $K$-means algorithm applied to the simulated data in the VGG16 latent space forms almost perfect clusters. Additionally, the VGG16+$K$-means approach finds high purity clusters of proton events for real experimental data. We also explore the application of clustering the latent space of autoencoder neural networks for event separation. While these networks show strong performance, they suffer from high variability in their results.

@@

The architecture is portrayed in Fig.~\ref{fig:mixae}, wherein tapered boxes denote a direction of compression in the network components. In the figure each encoder and decoder pair makes up an autoencoder. The assignment of the cluster belonging to a sample image, $\mathbf{\hat{x}} \in \mathcal{R}^{128 \times \hspace{0.01in} 128} $, is taken to be the index of the maximal element in the vector $\mathbf{p}$ as shown in the right-most part of the figure. Finally, the model is trained end-to-end with back propagation \cite{Linnainmaa1976}, as implemented in the machine learning package TensorFlow \cite{tensorflow}.

@@

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Filteredexamples.pdf}
\caption[Filtered proton samples by cluster belonging]{A sample of proton events from different $K$-means clusters from the filtered data set. The top row shows proton samples that are intermingled with noise events in the models' predictions, and the bottom row belongs to a cluster that contains proton events almost exclusively. Each row belongs to a single cluster corresponding to the filtered confusion matrix in Fig.~\ref{fig:clster_confmati_filt}}\label{fig:filtered_vgg_clster_repr}
\end{figure}

@@

\begin{figure}
\centering
	\includegraphics[width=\textwidth]{custom_work/Simulatedvgg_conf_mat.pdf} 
	\caption[Pre-trained network - confusion matrices]{Confusion matrix for the $K$-means clustering of simulated AT-TPC events. The true labels indicate samples belonging to the p (proton), or the carbon (C) class. We observe very high quality separation between the proton and carbon classes.}\label{fig:clster_confmat_sim}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\textwidth]{custom_work/Rawcache_conf_mat.pdf}
\caption[Pre-trained network - confusion matrices]{Confusion matrix for the $K$-means clustering of raw AT-TPC events. The true labels indicate samples belonging to the p (proton), carbon (C), or other classes. Each column denotes a cluster, with each cell in the column denoting the count of that rows' class in the cluster. We observe that cluster 0 is a high quality proton event cluster }\label{fig:clster_confmati_raw}
\end{figure}

>>>>

\begin{abstract}
% Modified by DB (6/8/2020)
%Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant. 
%on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. 
%On the semi-supervised task, we find that a logistic regression classifier trained on small labeled subsets of the latent space of these models perform very well, both on simulated and real data. On the clustering task, 
%With autoencoder neural networks we find improved descriptions of data from experiments.
This article presents novel applications of unsupervised machine learning methods to the problem of event separation in an active target detector, the Active-Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. 
The overarching goal is to group similar events in the early stages of the data analysis, thereby improving efficiency by limiting the computationally expensive processing of unnecessary events.
The application of unsupervised clustering algorithms to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on $^{46}$Ar is introduced. We explore the performance of autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network. We study clustering performance on data from a simulated $^{46}$Ar experiment, and records from the AT-TPC detector. 
We find that a $K$-means algorithm applied to simulated data in the VGG16 latent space forms almost perfect clusters. Additionally, the VGG16+$K$-means approach finds high purity clusters of proton events for real experimental data. We also explore the application of clustering the latent space of autoencoder neural networks for event separation. While these networks show strong performance, they suffer from high variability in their results.

@@

The architecture is portrayed in Fig.~\ref{fig:mixae}, wherein tapered boxes denote a direction of compression in the network components. In the figure each encoder and decoder pair makes up an autoencoder. The assignment of the cluster for a sample x-y event image, $\mathbf{\hat{x}} \in \mathcal{R}^{128 \times \hspace{0.01in} 128} $, is taken to be the index of the maximal element in the vector $\mathbf{p}$ as shown in the right-most part of the figure. Finally, the model is trained end-to-end with back propagation \cite{Linnainmaa1976}, as implemented in the machine learning package TensorFlow \cite{tensorflow}.

@@

\begin{figure}
\includegraphics[width=\textwidth]{custom_work/Filteredexamples.pdf}
\caption[Filtered proton samples by cluster belonging]{A sample of proton events from different $K$-means clusters from the filtered data set. The bottom row shows proton samples that are intermingled with noise events in the models' predictions, and the top row belongs to a cluster that contains proton events almost exclusively. Each row belongs to a single cluster corresponding to the filtered confusion matrix in Fig.~\ref{fig:clster_confmati_filt}}\label{fig:filtered_vgg_clster_repr}
\end{figure}

@@

\begin{figure}
\centering
	\includegraphics[width=\textwidth]{custom_work/Simulatedcache_conf_mat.pdf} 
	\caption[Pre-trained network - confusion matrices]{Confusion matrix for the $K$-means clustering of simulated AT-TPC events. The true labels indicate samples belonging to the p (proton), or the carbon (C) class. We observe very high quality separation between the proton and carbon classes.}\label{fig:clster_confmat_sim}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\textwidth]{custom_work/Rawcache_conf_mat.pdf}
\caption[Pre-trained network - confusion matrices]{Confusion matrix for the $K$-means clustering of raw AT-TPC events. The true labels indicate samples belonging to the p (proton), carbon (C), or other classes. Each column denotes a cluster, with each cell in the column denoting the count of that rows' class in the cluster. We observe that cluster 2 is a high quality proton event cluster }\label{fig:clster_confmati_raw}
\end{figure}
